---
title: "Assignment 3: Instrumental Variable Estimation"
subtitle: "Replicating the IV strategy in Stokes (2015)"
author: "EDS 241 / ESM 244 (DUE: 2/17/2026)"
format:
  html:
    #theme: sketchy
    css: styles.css
date: "February 10, 2026"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

------------------------------------------------------------------------

### Assignment instructions:

Working with classmates to troubleshoot code and concepts is encouraged. If you collaborate, list collaborators at the top of your submission.

All written responses must be written independently (in your own words).

Keep your work readable: Use clear headings and label plot elements thoughtfully (where applicable).

Submit both your rendered output and the Quarto (`.qmd`) file.

**Assignment submission (YOUR NAME)**: Vedika Shirtekar

------------------------------------------------------------------------

### Introduction

In this assignment, you will replicate the instrumental variable analysis from Stokes (2015), which examined how local wind turbine projects influenced electoral outcomes. Building on the matched dataset from Assignment 2, we will use Two-Stage Least Squares (2SLS) to estimate the causal effect of having a wind turbine proposed nearby on the change in Liberal Party vote share between 2007 and 2011. The instrument used in Stokes (2015) is a measure of local wind resource (average wind power, logged), which predicts where wind turbines are proposed. By using this instrument, we aim to isolate the portion of variation in turbine placement that is as-good-as-random, helping to meet the assumptions for causal identification.

**Study:** [Stokes, 2015 – Article](https://drive.google.com/file/d/1y2Okzjq2EA43AW5JzCvFS8ecLpeP6NKh/view?usp=drive_link)

**Data source:** [Dataverse – Stokes, 2015 replication data](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/SDUGCC)

------------------------------------------------------------------------

::: callout
Note: The estimates you obtain may not exactly match the published results in Stokes (2015) due to the alternative matching procedure used for processing the data in the previous assignment. Estimates should approximate the findings reported in *Table 2* of the article.
:::

------------------------------------------------------------------------

### Load packages

```{r}
library(tidyverse)
library(janitor)
library(here)
library(jtools)   # for export_summs (pretty regression tables)
library(AER)      # for ivreg (2SLS estimation)

```

------------------------------------------------------------------------

### Load the matched dataset (from Assignment 2)

> The matched_data has been preprocessed by matching on key covariates (e.g. pretreatment home values, education, income, population density) to improve balance between treated and control precincts. We will now use this data for the IV analysis. Make sure to re-code the `precinct_id` variable as a `factor`.

```{r}

matched_data <- read_csv("data/matched_data_subset.csv")

matched_data <- matched_data %>% mutate(precinct_id = as.factor(precinct_id),
                                       district_id = as.factor(district_id))
```

------------------------------------------------------------------------

### Part 1: IV Identification Rationale

Intuition for Using an Instrument:

**Question 1:** After matching on observables, why might we still need to utilize an instrumental variable approach to identify the causal effect of turbine proposals on vote share? In other words, what potential issues remain that an IV method can help address in this context? Use specific examples from the study to illustrate threats to a causal interpretation, then explain how an IV approach is designed to mitigate those threats.

*Response:* Matching on observables is often not sufficient enough to establish causality because it cannot account for unobserved factors that may influence both turbine placement and voting behavior. In Stoke's study, selection bias remains a primary concern even after matching similar precincts because developers may have chosen to propose turbines in areas where land was easier to obtain or where political opposition was expected to be weaker; there is selection bias in the project location. If these areas were also already experiencing declining support for the Liberal Party for unrelated/unobserved reasons, then a simple comparison could make a mistake by attributing the voter loss to the turbines. To address this issue, the study uses an instrumental variable (IV) approach that utilizes local wind resources. Because wind speed is determined by natural geography and is unrelated to political preferences, it can be used to isolate the portion of turbine placement that is random. Using a 2LS, the analysis first predicts turbine proposals based on wind speed and then estimates their effect on vote share. This approach allows the causal effect to be driven by geographic suitability for a turbine rather than political selection or other unobserved factors.

------------------------------------------------------------------------

### Part 2: Two-Stage Least Squares (2SLS) Step-Wise Implementation

#### 2A. First-Stage Estimation: Regress the treatment ($D$) on the instrument ($Z$)

$$D_i = \alpha_0 + \alpha_1 Z_i $$

a.  Estimate the first-stage regression of the treatment on the instrument (with controls). Regress `proposed_turbine_3km` on `log_wind_power`.
b.  Include the control variables used in Stokes (2015) for both stages: Distance to lakes, geographic coordinates (latitude & longitude) with their squares and interaction, plus district fixed effects.
c.  After running the first stage, report the F-statistic for the instrument.

```{r}
first_stage <- lm(proposed_turbine_3km ~ log_wind_power + mindistlake + mindistlake_sq +
                      latitude + longitude  + 
                      lat_sq + long_sq + long_lat + 
                      district_id, data = matched_data)

export_summs(first_stage, digits = 3,
             model.names = c("First stage: Propoposed Turbine 3km"),
             coefs = c("(Intercept)", "log_wind_power")) 

```

### Testing Instrument Relevance

Check instrument strength (F-statistic)

```{r}
print(paste("F-statistic:", summary(first_stage)$fstatistic[1]))
```

**Question 2A:** Based on the instrument relevance test reported in the study, would you conclude the instrument is strong enough to be credible? Explain what a weak instrument would mean in this setting: Specifically, what would it suggest about compliance with Ontario’s Green Energy Act policy?

*Response:* The first stage results indicate that the instrument is strong and credible with an F-statistic exceeding 10 (F = 14.71) produced from this model. The model tested here reports an F-statistic of 14.71 and the coefficient for wind power is statistically significant (p \< .05). The same is observed in the study with a reported F-statistic of 68 and a statistically significant coefficient for wind power (p \< .05). A weak instrument would indicate that wind availability does not strongly influence where turbines are proposed, suggesting that developers chose locations for other unobserved reasons, such as land access or lower political resistance. The high F-statistic indicates that turbine proposals are strongly driven by wind resources, which is consistent with the intent of the Green Energy Act in Ontario. As a result, the IV approach is able to isolate turbine placement that is independent of local political influences, thereby reducing concerns about strategic site selection by developers.

------------------------------------------------------------------------

#### 2B. Second Stage Estimation

#### Regress the outcome ($Y$) on the fitted values from the 1st stage ($\hat{X}_i$)

$$Y_i = \beta_0 + \beta_1 \hat{D}_i + \epsilon_i$$

a.  Now estimate the second stage of the 2SLS.
b.  First, use the first-stage model to generate the predicted values of proposed_turbine_3km for each precinct (these are $\hat{D}_i$).
c.  Add these predicted values as a new column in matched_data (e.g. proposed_turbine_3km_HAT).
d.  Then, regress the outcome change_liberal (the change in Liberal vote share from 2007 to 2011) on the predicted treatment (proposed_turbine_3km_HAT), including the same controls and fixed effects as in the first stage.
e.  Fill in the code for these steps below to obtain the second-stage regression results.

Save predicted values $\hat{X}_i$ from first stage

```{r}

matched_data$proposed_turbine_3km_HAT <- predict(first_stage)

```

#### Estimate the second-stage regression

$$LiberalVoteShare_i = \beta_0 + \beta_1 \widehat{ProposedTurbine}_i + ControlVariables... +  \epsilon_i$$

```{r}
second_stage <- lm(change_liberal ~ proposed_turbine_3km_HAT +
                       mindistlake + mindistlake_sq + 
                       latitude + longitude + 
                       lat_sq + long_sq + long_lat + district_id, data = matched_data)

export_summs(second_stage, digits = 3, 
             model.names = c("Second stage: Change in Liberal Vote Share"),
             coefs = c("(Intercept)", "proposed_turbine_3km_HAT") ) 
```

#### Interpreting the 2SLS Estimate

**Question 2B:** Imagine you are explaining your 2SLS findings to a policymaker in Ontario. What does the estimated coefficient on `proposed_turbine_3km_HAT` imply about the electoral impact of a local wind turbine proposal (within 3 km) on liberal vote share?

The estimated coefficient of -0.065 indicates that proposing a wind turbine within 3 km of a precinct led to a drop in support for the incumbent Liberal Party. Specifically, turbine proposals caused about a 6.5 percentage point decrease in the party's vote share compared to districts without a turbine proposal. Because this estimate is produced from the second stage of a 2SLS, it can be interpreted as a causal effect rather than a simple correlation between the predicted wind turbine within 3 km of a precinct and vote share of the incumbent party. Essentially, the decline in votes reflects voter backlash directly caused by the turbine projects/proposals and not by broader political trends.

**Question 2C:** Explain what it means that IV identifies a LATE in the context of the wind-turbine voting study. What specific subset of observations does the second-stage 2SLS estimate apply to, and what does it imply about interpretation and generalizability?

*Response:* In Stokes's wind turbine study, the LATE measures the causal impact of a turbine proposal only for the precincts where turbines were proposed because availability wind resources (ex. wind speed) were high. Simply put, the estimate applies to areas where turbine placement was driven by wind suitability and not to all precincts. It does not include precincts that would never receive turbines due to geographic and/or legal limits (ex. property or agricultural policy concerns).

This understanding matters for the interpretation and generalizability of the model results. The estimated 6.5 percentage drop in vote share reflects the voter response in precincts where turbine placement followed the Green Energy Act. While this provides a strong causal estimate for those areas, it may not apply everywhere. As a result, it cannot be assumed that the same electoral backlash would occur in areas with low wind or in other political contexts where both voter reactions and developer behavior for turbine projects differ.

------------------------------------------------------------------------

### Part 3: IV Assumptions and Validity

#### Evaluate Instrument Validity

**Question 3:** List the four key assumptions required for the IV strategy (2SLS) to identify a causal effect, and briefly explain what each one means in the context of this study. (Hint: think about what conditions a valid instrument must satisfy (relevance, exclusion,...)

*Response:*

-   Relevance assumption: The instrument should strongly predict the treatment. Here, wind resources must predict where turbines are proposed. Since windier areas are much more likely to receive wind turbine proposals (high F-statistic), this assumption is passed.

-   Exclusion restriction: The instrument should affect the outcome only through the treatment; the instrument should be exogenous (must be independent and not influenced by additional variables and outcome of the regression equation of interest). Here, wind speed should influence liberal vote share only by leading to turbine proposals and not through other factors, such as agricultural policy or politics in rural areas. This assumption would fail if wind directly affected voting behavior.

-   Exogeneity: The instrument cannot be related to unobserved political factors; the instrument must be independent and not influenced by additional variables + the outcome of the regression equation. Essentially, local wind speeds cannot be related to voters' established political preferences or additional socioeconomic traits.

-   Monotonicity or uniformity: The instrument must push the treatment in the same direction everywhere. In this case, higher wind power should always make turbine proposals more likely rather than less likely. Essentially, no precinct should be more likely to receive turbines because it is less windy.

------------------------------------------------------------------------

### Part 4: Estimate 2SLS using `AER::ivreg()`

📜 [SEE Documentation for specification details: AER package Viggnette Example](https://cran.r-project.org/web/packages/AER/vignettes/AER.pdf)

::: callout-tip
Syntax for specifying 2SLS using `ivreg()`:

`ivreg( Y ~ D + CONTROLS  | Z + CONTROLS , data )`

-   The first-stage predictor variables go after the `~` symbol
-   The second-stage predictor variables go after the `|` symbol
:::

```{r}
  
fit_2sls <- ivreg(change_liberal ~ proposed_turbine_3km + 
                      mindistlake + mindistlake_sq + 
                      latitude + longitude + 
                      lat_sq + long_sq + long_lat + 
                      district_id | 
            # right side with instrument (wind power) replacing treatment
                log_wind_power + 
                mindistlake + mindistlake_sq + 
                latitude + longitude + 
                lat_sq + long_sq + long_lat + 
                district_id, 
            data = matched_data
                      )

export_summs(fit_2sls, digits = 3,
             model.names = c("Change in Liberal Vote Share")
             #coefs = c("(Intercept)", "proposed_turbine_3km")
             ) 
```

#### Robustness checking strategies utilized in Stokes, 2015

**Question 4:** Choose two *robustness checks* from the paper that the authors use to increase confidence in their causal identification strategy. For each one, summarize the logic and findings from the robustness check in your own words:

*Response:* Testing for "strategic selection" using wind resources is an important robustness check used by Stokes because fixed effects models rely on the assumption of parallel trends; essentially, it is the assumption that treated and untreated precincts would have followed similar political paths in the absence of turbines. If developers placed turbines in areas already politically distancing themselves away from the Liberal Party, then the fixed effects estimates would be biased. To address this issue, Stokes uses an IV approach, where natural wind speed predicts turbine placement (Stokes, 2015, pg. 6). Because wind is determined more by geography rather than political strategy/affiliation, this creates an "as-if-random" source of variation in turbine placement (Stokes, pg. 6). The close alignment between the IV and fixed effects estimates suggests that the electoral/political penalty reflects a causal response to turbine placement and not due to pre-existing political shifts or strategic site selection. As a result, the IV approach utilizing wind speed as an instrument helps address selection bias in turbine location (Stokes, pg. 6).

Another robustness check Stokes considers is testing different distance buffers to establish the boundaries how "powerful" a causal effect can be detected away from a turbine, as well as to prevent spillover. Essentially, if wind turbines really caused the electoral backlash, then the political/electoral effect should be strongest for people living closest to them and weaken with distance. Stokes tests this by estimating the effect at different distance ranges (1 km, 2 km, 3 km, and beyond). If precincts far from turbines showed the same response as nearby ones, then it would suggest a broader regional trend rather than an effect for a local turbine. The results indicated that backlash is highly localized; vote share losses are significant up to 3 km and diminish beyond that distance (Stokes, pg. 11). This spatial pattern supports a causal interpretation of the response because it matches how turbines actually affect people (ex. nearby noise, visually displeasing, etc.) rather than through region-wide political shifts.

------------------------------------------------------------------------

### END

------------------------------------------------------------------------
